Does freshwater connectivity influence phosphorus retention in lakes?

### 2019-01-19 manuscript/manuscript.Rmd
fillin manuscript

---
title: A Template for Two or One-Column Vignettes

liftr:
  maintainer: "Joseph Stachelek"
  email: "stachel2@msu.edu"
  from: "rocker/geospatial:latest"
  pandoc: true
  sysdeps: 
    - grass
  cran: 
    - LAGOSNE
    - brms
    - dplyr
    - tidyr
    - modelr
    - data.tree
    - ggplot2
    - tidybayes
    - cowplot
  remotes: 
    - "jsta/nhdR"
    - "jsta/spnetwork"
    - "jsta/streamnet"
    - "jsta/rgrass7sf"

# Use letters for affiliations
author:
  - name: Joseph Stachelek
    affiliation: a
  - name: Patricia A. Soranno
    affiliation: a
address:
  - code: a
    address: Institute of Smoke and Magic, University of Sometown, Sometown, XY, 12345

@@

# Place eg a DOI URL or CRAN Package URL here
doi_footer: "https://cran.r-project.org/package=YourPackage"

# Abstract
abstract: |
  Your abstract will be typeset here, and used by default a visually distinctive font.
  An abstract should explain to the general reader the major contributions of the article.

# Optional: Acknowledgements
acknowledgements: |
  This template package builds upon, and extends, the work of the excellent
  [rticles](https://cran.r-project.org/package=rticles) package, and both packages rely on the
  [PNAS LaTeX](http://www.pnas.org/site/authors/latex.xhtml) macros. Both these sources are
  gratefully acknowledged as this work would not have been possible without them.  Our extensions
  are under the same respective licensing term
  ([GPL-3](https://www.gnu.org/licenses/gpl-3.0.en.html) and
  [LPPL (>= 1.3)](https://www.latex-project.org/lppl/)).

# Optional: One or more keywords
keywords:
  - one
  - two
  - optional
  - keywords
  - here

@@

# Customize footer, eg by referencing the vignette
footer_contents: "YourPackage Vignette"

@@

This *pinp is not PNAS* template started when the introduction to
[Rcpp](http://dirk.eddelbuettel.com/code/rcpp.html) by \cite{cardilleCarbonWaterCycling2007} 
was converted into this updated
[Rcpp Introduction](https://eddelbuettel.github.io/pinp/Rcpp-introduction.pdf)
vignette.  It is based on the
[pnas_article](https://github.com/rstudio/rticles/tree/master/inst/rmarkdown/templates/pnas_article)
template of the wonderful [rticles](https://cran.r-project.org/package=rticles) package by
. The conversion from markdown to latex is facilitated by
[rmarkdown](https://cran.r-project.org/package=rmarkdown)
 and [knitr](https://cran.r-project.org/package=knitr)
\citep{cardilleCarbonWaterCycling2007}. The underlying LaTeX macros are from
[pnas.org](http://www.pnas.org/site/authors/latex.xhtml).

The remainder of the document carries over from the corresponding
[pnas_article](https://github.com/rstudio/rticles/tree/master/inst/rmarkdown/templates/pnas_article)
template document. but has been edited and updated to our use case.  A
few specific tips follow.  In general, for fine-tuning some knowledge
of LaTeX is helpful.

## Author Affiliations

Per common academic best practice, you can include your department,
institution, and complete address, with the ZIP/postal code, for each
author. Use lower case letters to match authors with institutions, as
shown in the example. Authors with an ORCID ID may supply this
information at submission.

## Document Options

We support several options via the YAML header

- Setting a DOI or URL footer, for example for the CRAN package URL,
  which is placed in the bottom-left footer of the title page and even pages;
- Setting a footer label, for example _YourPackage Vignette_ stating
  your package, which is placed in the bottom-right footer on odd
  pages;
- Setting a free-form author field used on the inside footer;
- Optional _Draft_ watermarking;
- Line of custom text in subtitle (`date_subtitle`) suitable to give publication info of the draft, e.g. journal name in a post-print.

## References

Here we differ from PNAS and suggest natbib. References will appear in
author-year form. Use `\citet{}`, `\citep{}`, etc as usual.

We default to the `jss.bst` style. To switch to a different bibliography
style, please use `biblio-style: style` in the YAML header.

## Inline R Code

The PNAS sample included a fixed PNG image here, but this document prefers
to show the results and embedding of _R_ code.

```{r figex, fig.width=3, fig.height=3, cache=TRUE, echo=TRUE, fig.cap="Narrow ggplot2 figure"}
library(ggplot2)
ggplot(mtcars, aes(wt, mpg)) +
    geom_point(size=3, aes(colour=factor(cyl))) +
    theme(legend.position="none")
```

Here we use a standard knitr bloc with explicit options for

- figure width and height (`fig.width`, `fig.height`), both set to three inches;
- whether the code is shown (`echo=TRUE`); and
- the caption (`fig.cap`) as shown above.

## Digital Figures

Markdown, Pandoc and LaTeX support `.eps` and `.pdf` files.

Figures and Tables should be labelled and referenced in the standard way
using the `\label{}` and `\ref{}` commands.

The R examples above show how to insert a column-wide
figure. To insert a figure wider than one column, please use the
`\begin{figure*}...\end{figure*}` environment.

One (roundabout) way of doing this is to _not_ actually plot a figure, but to
save it in a file as the following segment shows:

```{r densityPlot, echo=TRUE}
library(ggplot2)
p <- ggplot(data = midwest,
            mapping = aes(x = area,
                          fill = state,
                          color = state)) +
    geom_density(alpha = 0.3)
## save to file
suppressMessages(ggsave("densities.pdf", p))
```

This file is then included via standard LaTeX commands.

\begin{figure*}
  \begin{center}
    \includegraphics[width=0.66\textwidth, height=3.5in]{densities} 
  \end{center}
  \caption{Wide ggplot2 figure}\label{fig}
\end{figure*}

## Typeset Code (But Do Not Run It)

We can also just show code.

```r
xx <- faithful[,"eruptions"]
fit <- density(xx)
plot(fit)
```

This simply used a pandoc bloc started and ended by three backticks,
with `r` as the language choice.  Similarly, _many_ other languages
can be typeset directly simply by relying on pandoc.

## Single column equations

Authors may use 1- or 2-column equations in their article, according to
their preference.

To allow an equation to span both columns, options are to use the
`\begin{figure*}...\end{figure*}` environment mentioned above for
figures, or to use the `\begin{widetext}...\end{widetext}` environment
as shown in equation \ref{eqn:example} below.

Please note that this option may run into problems with floats and
footnotes, as mentioned in the [cuted package
documentation](http://texdoc.net/pkg/cuted). In the case of problems
with footnotes, it may be possible to correct the situation using
commands `\footnotemark` and `\footnotetext`.

\begin{equation}
  \begin{aligned}
(x+y)^3&=(x+y)(x+y)^2\\
       &=(x+y)(x^2+2xy+y^2) \\
       &=x^3+3x^2y+3xy^3+x^3. 
       \label{eqn:example} 
  \end{aligned}
\end{equation}

<!-- pandoc writes all tables using longtable, which fails in 2-column mode

  Species                    CBS     CV     G3
  ----------------------- ------ ------ ------
  1\. Acetaldehyde           0.0    0.0    0.0
  2\. Vinyl alcohol          9.1    9.6   13.5
  3\. Hydroxyethylidene     50.8   51.2   54.0

  : Comparison of the fitted potential energy surfaces and ab initio
  benchmark electronic energy calculations

-->

>>>>

---
title: Does freshwater connectivity influence phosphorus retention in lakes?

liftr:
  maintainer: "Joseph Stachelek"
  email: "stachel2@msu.edu"
  from: "rocker/geospatial:latest"
  pandoc: true
  sysdeps: 
    - apt-utils
    - grass
  cran: 
    - LAGOSNE
    - brms
    - dplyr
    - tidyr
    - modelr
    - data.tree
    - ggplot2
    - tidybayes
    - cowplot
  remotes: 
    - "jsta/nhdR"
    - "jsta/spnetwork"
    - "jsta/streamnet"
    - "jsta/rgrass7sf"

# Use letters for affiliations
author:
  - name: Joseph Stachelek
    affiliation: a
  - name: Patricia A. Soranno
    affiliation: a
address:
  - code: a
    address: Department of Fisheries and Wildlife, Michigan State University, 480 Wilson Rd., East Lansing, MI 48824, USA

@@

# Place eg a DOI URL or CRAN Package URL here
doi_footer: "https://doi.org/ZenodoDOI"

# Abstract
abstract: |
  Lake water residence time and depth are known to be strong predictors of phosphorus (P) retention. However, there is substantial variation in P retention among lakes with the same depth and residence time. One potential explanatory factor for this variation is differences in freshwater connectivity of lakes (i.e. the type and amount of freshwater connections to a lake), which can influence watershed P trapping or the particulate load fraction of P delivered to lakes via stream connections. To examine the relationship between P retention and connectivity, we quantified several different measures of connectivity including those that reflect downstream transport of material (sediment, water, and nutrients) within lake-stream networks (lake-stream-based metrics) as well as those that reflect transport of material from hillslope and riparian areas adjacent to watershed stream networks (stream-based metrics). Because it is not always clear what spatial extent is appropriate for determining functional differences in connectivity among lakes, we compared connectivity metrics at two important spatial extents: the lake subwatershed extent and the lake watershed extent. We found that variation in P retention among lakes was more strongly associated with connectivity metrics measured at the broader lake watershed Fextent rather than metrics measured at the finer lake subwatershed extent. Our results suggest that both connectivity between lakes and streams as well as connectivity of lakes and their terrestrial watersheds influence P retention.

# Optional: Acknowledgements
acknowledgements: |
  This work was supported by the U.S. National Science Foundation’s Dynamics of Coupled Natural and Human Systems (CNH) e Program (award 1517823) and Macrosystems Biology Program (awards EF-1638679, EF-1638554, EF-1638539, and EF-1638550). PAS was also supported by the USDA National Institute of Food and Agriculture, Hatch project 1013544. We thank the following individuals for a friendly review of an earlier draft: K.S. Cheruvelil, I.M. McCullough, N. Smith, and A. Poisson.

# Optional: One or more keywords
keywords:
  - phosphorus
  - phosphorus retention
  - aquatic connectivity 
  - lake connectivity
  - lakes
  - Vollenweider model

@@

# Customize footer, eg by referencing the vignette
footer_contents: "Connectivity and phosphorus retention"

@@

<!-- Emails: JS, stachel2@msu.edu; PAS, soranno@msu.edu; -->
<!-- ORCIDS: -->
<!-- JS: https://orcid.org/0000-0002-5924-2464 -->
<!-- PAS: https://orcid.org/0000-0003-1668-9271 -->

Lake phosphorus (P) retention is an important characteristic of lakes that can be used to predict P concentrations and to evaluate lake sensitivity to nutrient O loading and eutrophication (Alexander et al. 2008; Milstead et al. 2013). Specifically, P retention is an integrated measure of internal P losses including permanent sedimentation, biological uptake, and other processes that remove P from the water column \citep{chapra2008surface}. P retention has been well studied in lakes because P determines lake trophic status and downstream watershed yields (i.e. export to terminal lakes and coastal estuaries). Although previous studies have shown that lake P retention is related to water residence time (Vollenweider 1975), large uncertainties exist around this relationship (Brett and Benjamin 2007, Milstead et al. 2013). These uncertainties can be large in lakes with intermediate water residence times, particularly compared to lakes with either extremely short or extremely long water residence times (Figure A1). For example, lakes with very long water residence times (on the order of a decade or longer) have complete or near complete P retention, while lakes with very short water residence times (on the order of days) have almost no P retention (Brett and Benjamin, 2007). The reason for the substantial uncertainty in lake P retention between these two extremes may be that predicting retention solely on the basis of water residence time does not capture many of the other factors and processes that affect P retention (Figure \ref{fig1}).

\begin{figure}
  \begin{center}
    \includegraphics[width=0.45\textwidth, height=3.5in,keepaspectratio]{../figures/01_conceptual_p-cycle_pt2} 
  \end{center}
  \caption{Major lake and watershed factors affecting lake P retention. Shaded symbols indicate factors typically considered in P retention models whereas open symbols indicate additional factors specific to the present study. Dashed lines indicate inferred relationships, which cannot be tested with available data, but are otherwise discussed herein.}\label{fig1}
\end{figure}

One well-studied factor that has been shown to influence lake P retention is lake depth through its influence on internal processing of P loads (Søndergaard et al. 2013; Vollenweider 1975). The mechanism for such an influence is that lake depth controls thermal stratification and material resuspension from the benthos. As a result, shallow lakes have a tendency to mix throughout the summer causing redistribution e of sedimented phosphorus throughout the mixed zone (Fee et al. 1996). This mixing and redistribution of sedimented P often leads to tighter benthic-pelagic coupling and increased P recycling e (Cha et al. 2013). Thus, depth is one example of a lake characteristic that is likely to influence P retention in concert with water residence time (Cheng et al. 2010).

A primary tool for evaluating the influence of specific lake characteristics, like lake depth, on P retention is statistical phosphorus retention modelling. Although the form of such models is variable, many models estimate P retention as a function of water residence time and a parameter k that represents in-lake P decay (Chapra 2008, Vollenweider 1975). Many studies treat k as a constant global parameter (i.e., it has the same value for all lakes), which may be valid only for studies that consider small numbers of lakes in a limited geographic region, with similar characteristics. Few studies have modeled different k values based on lake or watershed characteristics, despite the many differences among lakes that likely influence their ability to process P (Cheng et al. 2010).

For example, there is evidence that the relative proportion of particulate versus dissolved P loads (hereafter, particulate load fraction), influences P retention in stream and wetland ecosystems (Jarvie et al. 2011; Kronvang 1992; Russell et al. 1998; Vanni et al. 2001). However, evidence for particulate load fraction controls on P retention in lakes remains limited because of difficulties in tracking the fate of particulate loads after entering a lake (Brett and Benjamin 2007; Dillon and Molot 1996). Therefore, there is potential to further study particulate load Ffraction using proxies that may be closely related to it, such as the relative amounts of point and nonpoint nutrient sources to a lake subwatershed.

In general, point source inputs e to lakes are associated with increased dissolved P loads (Kronvang 1992; Russell et al. 1998), whereas non-point source inputs to lakes, such as those in lake watersheds with high agricultural land e use cover, typically have higher particulate P loads (Carpenter et al. 1998; Sharpley et al. 1994). Exceptions to these generalizations have been observed in areas where increased dissolved P loading O occurs not as a result of point source nutrient inputs but rather as a result of non-point source runoff due to P saturated soils or legacy P release (Bennett and Carpenter 2001; Powers et al. 2016). Despite the fact that particulate loads are usually related to non-point source inputs, the fraction of non-point source inputs that are in particulate form is highly variable and may depend on intra-annual flow variations as well as watershed erosion characteristics (Jarvie et al. 2011). Furthermore, quantifying non-point source inputs at broad spatial scales is not commonly done due to logistical and sampling constraints (Guy et al. 1994). As a result, studies often infer the relative amounts of particulate and dissolved loading from other available proxy data such as land-use cover (Ellison and Brett
2006, Djodjic and Markensten 2018).

Apart from land-use cover, another potential proxy for particulate load fraction entering lakes is the type and amount of freshwater connections to a lake, which we argue is easier to measure than other proxies, and could help to improve estimates of lake P retention, especially in lakes for which we lack P loading data (Figure 1). Although freshwater connectivity may be easy to measure from a logistical standpoint, there are still many ways to measure connectivity that likely represent different mechanisms of water and material flow (Figure 2). We broadly define and study two types of freshwater connectivity that correspond to either stream-based metrics or lake-stream-based metrics. First, lake-stream-based metrics measure the connections between a lake, other upstream lakes, and streams e in their watershed. This type of connectivity can be quantified by measuring the closest distance to an upstream stream-connected lake (Figure 2A), or by measuring the total upstream lake area e (Figure 2B). Second, stream-based metrics measure the connections between inflowing streams and their surrounding land whereby increasing stream connections lead to a greater abundance of land-water O interfaces and greater transport of material from hillslope and riparian areas adjacent to watershed stream networks (Figure 2C-E).

In addition to variation among connectivity metrics and connectivity metric types, it is also not always clear which spatial extent is appropriate for determining functional differences in connectivity among lakes (Soranno et al. 2015). Such information is needed to inform the design
of regulatory frameworks balancing controls on cumulative nutrient transport along stream networks and controls on localized nutrient transport (Alexander et al. 2008; Carpenter et al. 1998; Withers and Jarvie 2008). To test the importance of different spatial extents, we examined connectivity metrics measured for both the lake subwatershed extent and the lake watershed extent (Figure 3). Here, the lake subwatershed extent includes the elements of the immediate watershed in the direct drainage of a lake whereas the lake watershed extent includes all of the elements in the entirety of the lake-stream network up to and including headwater streams (Figure 3).
We propose that measures of freshwater connectivity are related to P retention in the following ways. First, some connectivity metrics reflect the proximity of terrestrial watershed areas to the stream network (Covino 2017). For example, lakes with a high watershed stream density should have increased particulate matter loading from terrestrial hillslope and riparian areas adjacent to watershed stream networks because particulate matter export is limited by overland distance (Gomi et al. 2002). Second, connectivity metrics can reflect the configuration of lakes within lake-stream networks. e For example, lakes with upstream lakes in close proximity may receive P loads that have previously undergone in-lake processing whereby labile fractions have already been trapped in upstream lakes e (Cardille et al. 2007). In contrast, lakes with more distant upstream lakes are more likely to receive the more labile fractions from terrestrial runoff that serve to increase P retention as opposed to receiving the more recalcitrant fractions that are resistant to biological uptake and are thus not retained. Although some connectivity metrics have an intuitive relation to P retention, it is not clear which specific measures of freshwater connectivity are important for transport of particulate matter. Therefore, our study is designed to examine and compare which measures of connectivity are more related to lake P retention. We addressed the above knowledge gaps by quantifying and comparing a range of freshwater connectivity measures at multiple spatial extents. Taken together, our suite of connectivity metrics reflect both freshwater connectivity in the direct drainage of lakes (i.e. the lake subwatershed extent) and freshwater connectivity of the entirety of the stream network up to and including watershed headwater areas (i.e. the lake watershed extent, Figure 3). Our motivation for measuring connectivity so many ways is that it is easy to measure connectivity of lakes with small watersheds situated at the beginning of lake chains but it is much more challenging to identify the type and extent of connectivity in larger, more complex lake networks. Another reason we examined relationships between P retention, multiple measures of connectivity, and multiple spatial extents is that many commonly used connectivity metrics merely reflect watershed size (spatial extent) rather than types of material transport or particulate load fraction (Leibowitz et al., 2018). Although lakes in larger watersheds have both a greater potential area from which to source particulate runoff and total phosphorus export from the watershed, we expect that delivery of sediment-bound phosphorus is dependent on connectivity-mediated trapping in the upstream watershed e (Prairie and Kalff 1986). We asked two questions in this  study: 1) Which measures of freshwater connectivity influence lake phosphorus retention? 2) What e spatial extent of connectivity most strongly influences P retention? To answer these questions, we fit statistical P retention models in a Bayesian hierarchical framework following Cheng et O al. (2010) where 2 separate values of the k processing parameter were estimated for lakes with either high or low values of each connectivity metric. Using this approach, higher k values for a specific lake connectivity class indicates more extensive in-lake processing and higher P retention. We applied this model to a dataset of 129 lakes across a wide range of hydrologic, geologic, and climatic settings. We fit separate models using each combination of connectivity metric and spatial extent in an effort to determine whether P retention is more strongly controlled at the lake subwatershed extent or the lake watershed extent.

## Methods

### Dataset description

We used data on P retention, maximum depth, and water residence time from 129 lakes in the National Eutrophication Survey (USEPA 1975, Stachelek et al. 2018). Mean annual P loading, P discharge, and P retention values in the National Eutrophication Survey (NES) dataset were calculated based on monthly sampling for P in tributary and outlet discharge points as well as
any municipal waste discharges from 1972 to 1975. Here, P retention is a unitless value representing the fraction of incoming P loads. Sampling frequency for water discharge and residence time varied among o
 lakes but details of these variations were not provided in the source
rdataset (USEPA 1975, Stachelek et al. 2018). Estimates of water discharge and residence time in
R
the NES dataset represent normalized e mean flow estimates expected to occur during a period of
average precipitation and hydrology v
 (USEPA 1975). Water residence time for our study lakes
iranged from 1 week to 17 years with an interquartile e range of 3 months to 1.8 years while P
retention ranged from 0.06 to 0.99 with an w
 interquartile range of 0.24 to 0.59 (Table 1).
We supplemented the NES dataset with boundaries O for lake subwatersheds, as well as
n
estimates of stream density, upstream lake area, upstream lake connection(s), baseflow (an index
ly
of groundwater inputs), land-use cover, and other water quality measurements from the LAGOS-
NE dataset (Table 1; Soranno et al. 2017). Our study lakes encompassed a wide range of land-
use cover types and nutrient levels (Table 1). Although, lake subwatersheds were variable with
respect to agricultural land use cover, we did not observe a strong relationship with lake P
retention (Figure A2). On average, the water quality (total phosphorus, chlorophyll
concentration, and Secchi depth) of the lakes in our study are similar to other US lakes as
measured by the stratified random sampling design of the National Lakes Assessment (NLA) lake population (USEPA 2016). However, our lakes are substantially larger and deeper than most
NLA lakes (Figure A3).
We restricted the lakes in the study to those located within the footprint of LAGOS-NE
which includes lakes located in 17 northeastern and midwestern US states (Soranno et al. 2017).
We excluded lakes from our analysis if they had a surface area of greater than 1000 km2 or a
surface area of less than 0.1 km2. We also excluded lakes if they had a maximum depth of greater
than 70 m, lacked upstream surface water connections, or had one of the North American Great
Lakes in its upstream watershed. A total of 129 out of 236 NES lakes met each of these selection criteria.

### Connectivity metrics and spatial extents

In addition to data from the NES and  LAGOS-NE datasets, we calculated several connectivity
imetrics that we expected would be related e to lake P retention (Figure 2). Some of these metrics
were stream-based with the goal of capturing w
 aspects of the configuration of each lakes’
upstream surface water network (Figure 2C-D). In particular, O we chose metrics that would
n
quantitatively approximate network complexity under the assumption that highly complex
ly
networks are also low connectivity networks. This assumption is supported by the findings of
stream network simulations where increased network complexity leads to increased network
resistance and ultimately decreases in network connectivity (Rodriguez-Iturbe and Rinaldo
2001). In addition to stream-based metrics, we calculated lake-stream-based metrics that we
expected would reflect the likelihood of P trapping in upstream lakes prior to arriving at a given
focal lake via tributary flow (Figure 2A, 2B). A simple metric that captures this likelihood is the
presence (or absence) of an upstream lake (greater than 4 ha) which we define as “lake

connection” (Fergus et al. 2017). In addition to lake connection, we calculated related metrics
such as total upstream lake area and the network distance to the closest upstream lake.
To examine the importance of spatial extent relative to our connectivity metrics, we
calculated each metric at multiple extents (Figure 3). First, we calculated connectivity metrics at
the scale of individual lake subwatersheds. We defined a lake subwatershed as the area draining
into a particular lake exclusive of any upstream areas that drain into a lake greater than or equal
to 10 ha (0.1 km2). Next, we calculated connectivity metrics at the scale of entire upstream lake
networks (lake watershed extent). We defined a lake watershed as the area draining into any part
Fof the upstream network o
 irrespective of the presence or absence of upstream lakes (Figure 3).
rAll connectivity metrics were calculated using the high-resolution National Hydrography
R
Dataset (NHD) as a primary input (USGS e 2018). Average link length was calculated as the total
stream length in a given watershed divided v
 by the number of stream reaches after dissolving
i(removing) any network points that do not e occur at a stream junction. Stream density was
calculated as the length of all streams in the w
 watershed (minus artificial lines through lakes)
expressed in units of meters per hectare. Upstream lake O area was calculated as the sum of the
n
lake area in the upstream watershed expressed in square meters. Stream order ratio was defined
ly
as the number of headwater (first-order) streams in the upstream watershed of the focal lake
divided by the total number of higher order (> 1) streams (Barbera and Rosso 1989). Closest
distance to an upstream lake was defined as the shortest path-distance (rather than the straight-
line distance) to a lake upstream from the focal lake.
We calculated all connectivity metrics and lake watershed extents using the streamnet
and nhdR R packages respectively (Stachelek 2018a, Stachelek 2018b). The algorithms in the
streamnet package use the sf R package (Pebesma 2018) as well as the v.net and v.stream.order modules (Jasiewicz and Metz 2011) included in GRASS GIS (GRASS Development Team
2017). All processed connectivity data and code are available at [Zenodo DOI].

### Modelling lake P retention

We modelled lake total phosphorus retention (hereafter, P retention) using the Vollenweider equation that models P retention as a function of water residence time and a parameter (k)
conceptually representing in-lake P decay (Chapra 2008, Vollenweider 1975). Although there are
several variants of this basic equation, we selected a 2-parameter form (equation 1) that has been
Fshown to have good performance o
 in multiple cross-sectional studies (Brett and Benjamin 2007,
rCheng et al. 2010):
R
Ri = 1 ― 1 + 1
 ktx
 i
 e Eq. 1
v
where Ri is P retention as a fraction of P inputs, i $\tau$ is water residence time, k is a unitless
eparameter representing in-lake P decay, and w
 x is a unitless parameter representing P export via
hydrologic flushing. Here, higher values of k are associated with greater integrated P losses from
Osedimentation and biological uptake resulting in greater P retention. Note that Eq 1 does not
n
include a recycling term. Therefore, our results represent net l P retention (as opposed to gross P
y
retention) under a steady state assumption where lakes are at equilibrium with respect to
recycling (Vollenweider 1975). Note that although some forms of the Vollenweider equation use
P loading as a predictor variable, it does not appear in Eq 1. The reason for this is two-fold. First,
estimates of P loading are more difficult to obtain than estimates of water residence time and our
aim was to develop a model than can be widely applied to lakes for which we lack detailed data
on P loading. Second, loading based model forms have been shown to be mathematically
equivalent to water residence time based model forms (Brett and Benjamin 2008).
We used the model described by Eq 1 to compare lake P retention in lakes with different
connectivity by fitting the model in two ways. First, we modelled the overall relationship
between P retention and water residence time for all lakes in our dataset (global model). Second,
we fit hierarchical versions of Eq. 1 where k was modelled separately (kj) as a function of a
binary sub-population indicator gi denoting membership in one of two lake classes formed on the
basis of specific connectivity metrics (or lake depth) and specific spatial extents:
1
Ri = 1 ―
 1 + kjtxi
 F Eq. 2
o
kj = gi
 rwhere greater differences in k between R
 the two groups indicate greater support for a connectivity
eeffect on P retention. Prior to model fitting, we examined the bivariate relationships between
v
each connectivity metric, water residence i time, P loading, and other factors related to P retention
eusing Pearson’s correlation coefficients. The w
 purpose of this exercise was twofold, to determine
the potential for collinearity among any of the variables in Eq 2 and to identify any relationships
Obetween P retention, water residence time, and other watershed and lake characteristics that were
n
not included in our model. As only one connectivity metric was l used to define g for each model
y
we did not use the results of this exercise to exclude variables from further investigation. We
quantified the relative support for an effect of each connectivity metric on P retention in more
detail by calculating the difference in the median value of the P decay parameter k between
groups (i.e. $\delta$k). We used these median k values along with median estimates of x to determine
how differences in k translated to differences in P retention (Eq 2). We judged significance by
whether or not differences in group-wise P retention were greater than the measurement
precision of P retention (> 0.01).

We fit all models in a Bayesian framework using the non-linear extension to the brms
package to interface with the Stan statistical program (Burkner 2017, Stan Development Team
2017). In both models, we set a semi-informative prior on k and x of N(1.3, 0.1) and N(0.45, 0.1)
respectively. These priors were based on the confidence intervals presented in Brett and
Benjamin (2007) and qualitatively matched those used by Cheng et al. (2010). We used the
default settings of brms and rstan to generate posterior estimates using four chains of 4,000
iterations each with no thinning and initial parameter values drawn from a uniform distribution
bounded between -2 and 2. We also used the brms package for model evaluation by computing a
FBayesian R2 following o
 the method of Gelman et al. (2017).

### Lake connectivity classes

Our lake connectivity classes were formed by dividing the lake dataset into two classes for each
iconnectivity metric based on the bivariate e relationship between each metric and P retention. Prior
studies have used a similar binary splitting w
 approach to examine the effect of various exogenous
factors on lake P retention (Cheng et al. 2010; Shimoda O and Arhonditsis 2015). We determined
n
splitting criteria for each metric from the results of a random forest procedure incorporating
ly
conditional inference trees (ctree, Hothorn and Zeileis 2015). This procedure creates binary splits
of the independent variables (i.e. each of the connectivity metrics), which are recursively
repeated to find the split that maximizes association with the dependent variable (P retention).
The advantage of the ctree technique over the more typical classification-regression tree (CART)
technique is that tree growth stopping rules are pre-specified (Hothorn et al., 2006). As a result,
some of subjectivity associated with post-hoc tree pruning is avoided.

## Results

### Interactions between connectivity, hydrology, and P loading

We examined the bivariate relationships between connectivity, water residence time, P loading and other factors related to P retention to determine the potential for strong relationships among
any of the variables which were not accounted for by our model structure (Eq 2, Figure 4). We
found evidence for some relationships among these variables, but none that suggest either
redundancy among connectivity metrics or that otherwise limit our ability to infer relationships
with P retention (Figure 4). For example, the Pearson correlation (r) between water residence
Ftime and stream density, o
 which is implicitly accounted for in our model structure, was 0.29 (p <
r0.05). In contrast, the correlation between water residence time and upstream lake area, which is
R
unaccounted for in our model structure, e was only 0.10 (p > 0.05). Note that interactions between
connectivity metrics and water residence v
 time are accounted for in our hierarchical model
istructure because Eq. 2 uses a global coefficient e x for water residence time that is estimated
separately from the hierarchical and connectivity-dependent w
 P decay coefficient k. This is
conceptually similar to fitting the k-connectivity-P retention O relationship to the residuals of the
n
water residence time to P retention relationship. Perhaps surprisingly, we did not observe strong
ly
correlations between P loading and water residence time (r = 0.13, p > 0.05) or between lake
depth and most connectivity metrics (r < 0.17). This suggests that our P retention results are not
confounded by variations in lake depth, by interactions between connectivity and P loading, or
by interactions between water residence time and P loading.
A secondary purpose of examining the bivariate relationships between connectivity,
water residence time, P loading and other factors related to P retention was to determine if our
connectivity metrics were related to each other such that they provide similar information. With the exception of stream density and baseflow (r = -0.52, p < 0.05), correlations among
connectivity metrics were low and of a similar magnitude as the correlations between
connectivity metrics and water residence time (0.10 < r < 0.29). The strongest correlation
between any connectivity metric or lake characteristic was upstream lake area and lake
watershed area (r = 0.77, p < 0.05).
We observed notably strong correlations between P loading and several lake
characteristics not accounted for in Eq 2. These include the correlation between P loading and
lake watershed area (r = 0.62, p < 0.05) as well as the correlation between P loading and lake
Fsubwatershed area (r = o
 0.76, p < 0.05). In addition, the correlation between P loading and
rupstream lake area was quite strong (r = 0.52, p < 0.05). Despite strong correlations between P
R
loading, watershed area, and upstream e lake area, we did not observe a strong correlation between
P loading and P retention (r = 0.15, p > 0.05). Notably, this correlation was much weaker than the correlation between water residence time e and P retention (r = 0.44, p > 0.05). Our
observation that the correlation between P w
 retention and P loading was not appreciably stronger
than correlations between P retention and our connectivity O metrics suggests that our estimates of
n
connectivity metric effects on P retention are not confounded by interactions between water
ly
residence time and P loading.
Although we did not observe strong correlations among connectivity metrics, we found
that lakes with similar connectivity metric values, were spatially clustered (Figure 5). In
particular, we found that lakes were concentrated in either the southern or northern portions of
our study area depending on their connectivity metric value (Figure 5). This observation is
consistent with the findings of Fergus et al. (2017) that lakes in the northern portion of our study area have distinct freshwater connectivity as compared to lakes in the southern portion of our
study area.

### Effects of connectivity on P retention

We found that freshwater connectivity metrics were associated with lake P retention (Figure 6,
7). Across all connectivity metrics except stream order ratio, we found that the P decay
coefficient k and thus P retention was associated with whether a lake had a high or low value of
each connectivity metric. These findings matched our expectations in several ways. Most
Fnotably, lakes with shorter o
 average link lengths had higher P retention relative to lakes with
rlonger average link lengths. In addition, lakes with less upstream lake area had higher P retention
R
than lakes with more upstream lake e area (Figure 6, 7).
We found that some connectivity v
 metrics were more strongly related to P retention than
iothers. For example, the model R2 for network e average link length was higher (R2 = 0.41) than
the global model (R2 = 0.34). For other connectivity w
 metrics, such as network stream order ratio,
goodness-of-fit (R2 = 0.36) was very similar to the global O model (R2 = 0.34). Model fit for other
n
connectivity metrics was in between these two extremes (0. 36 < R2 < 0.41). Overall, we found
ly
that all hierarchical models had at least a marginally better fit to the water residence time versus
P retention relationship than a global model which does not account for connectivity (Figure 6).
Although we found a discernable effect of connectivity metrics on lake P retention, the
somewhat modest improvements in model fit may be due to the fact that water residence time
remains a dominant effect even after accounting for freshwater connectivity.

### Comparison across connectivity metrics and spatial extent

Differences in P retention among lakes with different connectivity metric values was reflected in
differences among connectivity class-specific values of the P decay parameter k (Figure 7). The
connectivity metric that had the most effect on k was average link length (Table 2). For instance,
hierarchical models fit with the average link length metric had a median effect size of 0.23 and
0.05 for k and P retention respectively (Table 2), which means that for this metric, lakes with
shorter average link lengths retained 4.7 to 4.9 % more P than lakes with longer average link
lengths. The influence of lake-stream-based connectivity metrics on lake P retention was similar
to stream-based connectivity metrics (Figure 7). This suggests that both lake-stream-based
Fconnectivity between lakes o
 and streams as well as stream-based connectivity of lakes and their
rterrestrial watersheds influence P retention (Figure 2).
R
We found that connectivity metrics e measured at the lake watershed extent were more
strongly associated with P retention v
 than metrics measured at the lake subwatershed extent
i(Table 2, Figure 7). Specifically, the metrics e that had the strongest association with P retention
such as average link length ($\delta$k = 0.23), closest w
 lake distance ($\delta$k = 0.22), and stream density ($\delta$k
= 0.20) also had a stronger association at the lake watershed O extent rather than at the lake
n
subwatershed extent (Table 2, Figure 7). An exception to this pattern was observed with the
ly
baseflow connectivity metric where although greater differences at the lake subwatershed extent
were more strongly associated with P retention, the sign of the effect was variable depending on
measurement extent (i.e. the value of the P decay parameter was positively related to
connectivity metric values at the subwatershed extent but was negatively related at the watershed
extent). For several connectivity metrics, we judged that differences in P decay among lakes with
either low or high connectivity metric values were not significant because they translated to
differences in P retention that were less than measurement precision (Table 2).

## Discussion
Although prior studies have found that lake P retention is related to water residence time (Brett
and Benjamin, Cheng, et al. 2010, Vollenweider 1975), there is substantial variation around this
relationship, particularly at intermediate water residence times. We found that some of this
remaining variation could be explained using a hierarchical modelling framework that accounts
for differences in freshwater connectivity among lakes. Although we found that the magnitude of
this effect depends on the specific connectivity metric, our results are consistent with the findings of previous studies o
 showing that connectivity metrics are associated with differences in
rlake carbon input fluxes and differences in lake nitrogen output fluxes (Cardille et al. 2007;
R
Schmadel et al. 2018). We also found e important differences in the association between
connectivity metrics and P retention v
 at different measurement extents. Specifically, we found
ithat P retention was more strongly associated e with connectivity measured at the broader lake
watershed extent rather than connectivity measured w
 at the finer lake subwatershed extent.

### Connectivity and P retention
